Learning how to act: making good decisions with machine learning

### 2017-09-13 thesis/thesis_map.tex
fixes references

Causality is often liked to explanation; understanding how and why things happen. I view explanation in terms of compression and generalisabilty: the amount of information about the world a model can capture. This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction. A standard predictive model encodes all the information needed to predict some output given inputs provided the system generating the data does not change. A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed. More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts. By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}. Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.

\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{what if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs. The former aims to identify the effect of a known cause. The latter can be viewed as identifying causes of an effect. They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, that we describe in section \ref{sec:causal_models}. However, they state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined. The goal of identifying the causes of an effect can be formalised within the graphical causal model framework. Solving this problem is certainly much more challenging than identifying the effect of a specific intervention on a given outcome, since it requires us to test or infer the effect of interventions on many different variables. These practical difficulties may well be overwhelming, particularly in fields such as social science and economics where data sets are often relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "what if" questions are hard to resolve conclusively. However, this does not mean that the problem of identifying causes of effects is ill-posed in principle. It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich literature, see for example \citep{}.
\todo{reference a review paper on causal discovery here}

@@

Playing Go (and other games) is a case with some subtleties. At every turn, the AI agent has a number of actions available. The state of the board following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win \footnote{This is a challenging pattern recognition problem. There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win. It must identify higher level features of the board state that are associated with winning.}. Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of a player selecting action $a$, given board state $s$. This allows the agent to estimate the likelihood of winning from a given starting state by simulating many times the remainder of the game, drawing actions from $\P{a|s}$ for both players. Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a combination of these approaches \citep{silver2016mastering}. The supervised learning was enhanced by having the agent play (variants) of itself many times, so that its estimates of value for each board state and of the likelihood an opponent will play a given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X \text{ (Treatment)}$};
\node[node, right=of 2](3){$Y \text{ (Outcome)}$};
\path[]
	(2) edge (3);
\end{tikzpicture}
\caption{Causal Bayesian network for example \ref{exm:ranomized_experiment}}
\label{fig:causal_network_example}
\end{figure}

@@

A causal Bayesian network represents much more information than a Bayesian network with identical structure. A causal network encodes all possible interventions that could be specified with the do-notation. For example, if the network in figure \ref{fig:causal_adjust} were an ordinary Bayesian network and all the variables were binary, the associated distribution could be described by seven parameters. The equivalent causal Bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions. Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional thirty parameters \footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters. After each two variable intervention, we have a distribution over one variables which requires one parameter. This takes us to a total of $6\times3+12\times1 = 30$ additional parameters.}.

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Prescribed Treatment};
\node[node, below right=of 1](2){Actual Treatment};
\node[node, right=of 2](3){Outcome};
\node[lt, above left=of 3](4){U};
\path[]
	(1) edge (2)
	(2) edge (3) 
	(4) edge (3) edge (2);
\end{tikzpicture}
\caption{Randomised treatment with imperfect compliance}
\label{fig:randomized_imperfect_compliance}
\end{figure}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X = f_x(\epsilon_x)$};
\node[node, right=of 2](3){$Y = f_y(X,\epsilon_y)$};
\node[lt, above=of 3](4){$\epsilon_y$};
\node[lt, above=of 2](5){$\epsilon_x$};
\path[]
	(2) edge (3)
	(4) edge (3)
	(5) edge (2);
\end{tikzpicture}
\caption{SEM for example \ref{exm:ranomized_experiment}}
\label{fig:sem_randomized_treatment}
\end{figure}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X = f_x(Z,\epsilon_x)$};
\node[node,above right=of 2](6){$Z = f_z(\epsilon_z)$};
\node[lt, above=of 6](7){$\epsilon_z$};
\node[node, below right=of 6](3){$Y = f_y(X,Z,\epsilon_y)$};
\node[lt, above=of 3](4){$\epsilon_y$};
\node[lt, above=of 2](5){$\epsilon_x$};
\path[]
	(2) edge (3)
	(4) edge (3)
	(5) edge (2)
	(7) edge (6)
	(6) edge (2) edge (3);
\end{tikzpicture}
\caption{SEM for example \ref{exm:adjusting}}
\label{fig:sem:preschool}
\end{figure}

@@

Clarify Cheng's Questions - is there a mechanical method to go from one framework to another, does it matter if variables are discrete or continuous. \todo{}

@@

\begin{table}[h]
\caption{Treatment results}
\label{tab:simpson1}
\center
\begin{tabular}{c c c c c}
Treatment & Success & Fail & Total & Success Rate\\
\hline
\rowcolor[gray]{0.9} 
A &  87 & 13 & 100 & 87\%\\
\rowcolor{white} B & 75 & 25 & 100 & 75\%\\
\hline
\end{tabular}
\end{table}

@@

\begin{table}[h]
\caption{Treatment results by gender}
\label{tab:simpson2}
\center
\begin{tabular}{c c c c c c}
Gender & Treatment & Success & Fail & Total & Success Rate\\
\hline
\rowcolor[gray]{0.9}
M & A & 12 & 8 & 20 & 60\%\\
M & B & 56 & 24 & 80 & 70\%\\
\rowcolor[gray]{0.9}
F & A & 75 & 5 & 80 & 94\%\\
F & B & 19 & 1 & 20 & 95\%\\
\hline
\end{tabular}
\end{table}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Z = Gender};
\node[node, below left=of 1](2){T = Treatment};
\node[node, below right=of 1](3){Y = Outcome};
\path[]
	(1) edge (2) edge (3) 
	(2) edge (3);
\end{tikzpicture}
\caption{An example of causal network that can give rise to Simpson's Paradox. In this case, we should select treatment on the basis of the gender-specific results.}
\label{fig:simpsons_paradox}
\end{figure}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Z = Blood Pressure};
\node[node, below left=of 1](2){T = Treatment};
\node[node, below right=of 1](3){Y = Outcome};
\path[]
	(1) edge (3) 
	(2) edge (3)
	(2) edge (1);
\end{tikzpicture}
\caption{Another causal network that can exhibit Simpson's paradox. In this case, \quotes{the solution} is not to stratify on $Z$.}
\label{fig:simpsons_paradox2}
\end{figure}

@@

The structure in figure \ref{fig:dseptripple4} is referred to as a collider or v-structure. The counter-intuitive result that conditioning on $Z$ introduces dependence between $X$ and $Y$ is called the \textit{explaining away phenomena}. As an example, consider a scholarship available to female or disadvantaged students. Let $X$ be gender, $Y$ be family background and $Z$ receipt of the scholarship. There are roughly equal numbers of boys and girls in both poor and wealthy families so $X$ and $Y$ are independent. However, if we know a student is receiving a scholarship, then learning that they are male  increases the probability that they are disadvantaged.
\begin{figure}[h]
\centering
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y| Z$}
\label{fig:dseptripple1}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$X$};
\node[main node, below =of 1](2){$Z$};
\node[main node, below =of 2](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    (2) edge node {} (3);	
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y|Z$}
\label{fig:dseptripple2}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$Y$};
\node[main node, below =of 1](2){$Z$};
\node[main node, below =of 2](3){$X$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    (2) edge node {} (3);	
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y|Z$}
\label{fig:dseptripple3}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$Z$};
\node[main node, below left =of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
        edge node {} (3);	
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y|\{\}$}
\label{fig:dseptripple4}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$Z$};
\node[main node, above left =of 1](2){$X$};
\node[main node, above right =of 1](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge node {} (1)
    (3) edge node {} (1);	
\end{tikzpicture}
\end{subfigure}
\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
\label{fig:dseptriples}
\end{figure}

@@

\begin{figure}[h]
\centering
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{original network, $G$}
\label{fig:mutilatedOriginal}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]

@@

\begin{figure}[h]
\centering
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{original network, $G$}
\label{fig:rule2_orig}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]

@@

\begin{figure}[h]
\centering
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{original network, $G$}
\label{fig:rule22_orig}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]

@@

\begin{figure}[h]
\centering
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{original network, $G$}
\label{fig:rule3_original}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]
 %nodes
\node[main node](1){$Z$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, below right=of 2](4){$W$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (4)
    (4) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{augmented graph $G^{\hat{X}}$}
\label{fig:rule3_augmented}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]
 %nodes
\node[main node](1){$Z$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, below right=of 2](4){$W$};
\node[main node, above left=of 2](5){$\hat{X}$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (4)
    (4) edge (3)
    (5) edge (2);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{$G_{\overline{X}}$}
\label{fig:rule3_after}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]
 %nodes
\node[main node](1){$Z$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, below right=of 2](4){$W$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (3)
    (2) edge (4)
    (4) edge (3);
\end{tikzpicture} 
\end{subfigure}
\caption{Example application of equation \ref{eq:Do3}. $(Y \ci \hat{X}|W,Z) \implies \P{Y|do(X),W,Z} = \P{Y|W,Z}$. We have to condition on $Z$ because conditioning on $W$ blocks the path $\hat{X}-X-W-Y$ but opens $\hat{X}-X-Z-Y$.}
\label{fig:rule3}
\end{figure}

@@

\citet{Johansson2016_counterfactual_rep} identified that this is equivalent to the covariate shift problem. If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$, but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$. \footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift. Take the case where we have a binary intervention $x \in \set{0,1}$. Suppose we learn $h(1,\vb{z}) = \E{Y|x=1,\vb{z}}+g(\vb{z})$ and $h(0,\vb{z}) = \E{Y|x=0,\vb{z}}+g(\vb{z})$, then the estimated average causal effect equals the true average causal effect for any function $g$, $\E{h(1,\vb{z})-h(0,\vb{z})} = \E{Y|x=1,\vb{z}} - \E{Y|x=0,\vb{z}}$. More generally, if the goal is to select an optimal action $x^*$ from a continuous space of possible interventions we need algorithms capable of leveraging any structure in the relationship between $x$ and $y$ as well as a means of focusing the loss on regions of the sample likely to affect $x^*$.} With this connection to covariate shift in mind, let us return to regression, matching and propensity scores.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
\caption{
Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data. In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities. If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$. We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points). In both cases the linear model is a good fit to the data. However, the resulting estimate of the causal effect is very poor for the lower values of $z$. } 
\label{fig:covariate_shift_adjust}
\end{figure}

>>>>

Causality is often liked to explanation; understanding how and why things happen. I view explanation in terms of compression and generalisabilty: the amount of information about the world a model can capture. This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction. A standard predictive model encodes all the information needed to predict some output given inputs provided the system generating the data does not change. A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed. More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts. By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}. Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}

\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{what if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs. The former aims to identify the effect of a known cause. The latter can be viewed as identifying causes of an effect. They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, that we describe in section \ref{sec:causal_models}. However, they state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined. The goal of identifying the causes of an effect can be formalised within the graphical causal model framework. Solving this problem is certainly much more challenging than identifying the effect of a specific intervention on a given outcome, since it requires us to test or infer the effect of interventions on many different variables. These practical difficulties may well be overwhelming, particularly in fields such as social science and economics where data sets are often relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "what if" questions are hard to resolve conclusively. However, this does not mean that the problem of identifying causes of effects is ill-posed in principle. It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich literature, see \citet{Spirtes2016} for a recent review.

@@

Playing Go (and other games) is a case with some subtleties. At every turn, the AI agent has a number of actions available. The state of the board following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win.\footnote{This is a challenging pattern recognition problem. There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win. It must identify higher level features of the board state that are associated with winning.} Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of a player selecting action $a$, given board state $s$. This allows the agent to estimate the likelihood of winning from a given starting state by simulating many times the remainder of the game, drawing actions from $\P{a|s}$ for both players. Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a combination of these approaches \citep{silver2016mastering}. The supervised learning was enhanced by having the agent play (variants) of itself many times, so that its estimates of value for each board state and of the likelihood an opponent will play a given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.

@@

\begin{figure}[ht]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]

@@

\begin{figure}[ht]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X \text{ (Treatment)}$};
\node[node, right=of 2](3){$Y \text{ (Outcome)}$};
\path[]
	(2) edge (3);
\end{tikzpicture}
\caption{Causal Bayesian network for example \ref{exm:ranomized_experiment}}
\label{fig:causal_network_example}
\end{figure}

@@

A causal Bayesian network represents much more information than a Bayesian network with identical structure. A causal network encodes all possible interventions that could be specified with the do-notation. For example, if the network in figure \ref{fig:causal_adjust} were an ordinary Bayesian network and all the variables were binary, the associated distribution could be described by seven parameters. The equivalent causal Bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions. Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional thirty parameters.\footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters. After each two variable intervention, we have a distribution over one variables which requires one parameter. This takes us to a total of $6\times3+12\times1 = 30$ additional parameters.}

@@

\begin{figure}[ht]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Prescribed Treatment};
\node[node, below right=of 1](2){Actual Treatment};
\node[node, right=of 2](3){Outcome};
\node[lt, above left=of 3](4){U};
\path[]
	(1) edge (2)
	(2) edge (3) 
	(4) edge (3) edge (2);
\end{tikzpicture}
\caption{Randomised treatment with imperfect compliance}
\label{fig:randomized_imperfect_compliance}
\end{figure}

@@

\begin{figure}[ht]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X = f_x(\epsilon_x)$};
\node[node, right=of 2](3){$Y = f_y(X,\epsilon_y)$};
\node[lt, above=of 3](4){$\epsilon_y$};
\node[lt, above=of 2](5){$\epsilon_x$};
\path[]
	(2) edge (3)
	(4) edge (3)
	(5) edge (2);
\end{tikzpicture}
\caption{SEM for example \ref{exm:ranomized_experiment}}
\label{fig:sem_randomized_treatment}
\end{figure}

@@

\begin{figure}[ht]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X = f_x(Z,\epsilon_x)$};
\node[node,above right=of 2](6){$Z = f_z(\epsilon_z)$};
\node[lt, above=of 6](7){$\epsilon_z$};
\node[node, below right=of 6](3){$Y = f_y(X,Z,\epsilon_y)$};
\node[lt, above=of 3](4){$\epsilon_y$};
\node[lt, above=of 2](5){$\epsilon_x$};
\path[]
	(2) edge (3)
	(4) edge (3)
	(5) edge (2)
	(7) edge (6)
	(6) edge (2) edge (3);
\end{tikzpicture}
\caption{SEM for example \ref{exm:adjusting}}
\label{fig:sem:preschool}
\end{figure}

@@

@@

\begin{table}[ht]
\caption{Treatment results}
\label{tab:simpson1}
\center
\begin{tabular}{c c c c c}
Treatment & Success & Fail & Total & Success Rate\\
\hline
\rowcolor[gray]{0.9} 
A &  87 & 13 & 100 & 87\%\\
\rowcolor{white} B & 75 & 25 & 100 & 75\%\\
\hline
\end{tabular}
\end{table}

@@

\begin{table}[ht]
\caption{Treatment results by gender}
\label{tab:simpson2}
\center
\begin{tabular}{c c c c c c}
Gender & Treatment & Success & Fail & Total & Success Rate\\
\hline
\rowcolor[gray]{0.9}
M & A & 12 & 8 & 20 & 60\%\\
M & B & 56 & 24 & 80 & 70\%\\
\rowcolor[gray]{0.9}
F & A & 75 & 5 & 80 & 94\%\\
F & B & 19 & 1 & 20 & 95\%\\
\hline
\end{tabular}
\end{table}

@@

\begin{figure}[ht]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Z = Gender};
\node[node, below left=of 1](2){T = Treatment};
\node[node, below right=of 1](3){Y = Outcome};
\path[]
	(1) edge (2) edge (3) 
	(2) edge (3);
\end{tikzpicture}
\caption{An example of causal network that can give rise to Simpson's Paradox. In this case, we should select treatment on the basis of the gender-specific results.}
\label{fig:simpsons_paradox}
\end{figure}

@@

\begin{figure}[ht]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Z = Blood Pressure};
\node[node, below left=of 1](2){T = Treatment};
\node[node, below right=of 1](3){Y = Outcome};
\path[]
	(1) edge (3) 
	(2) edge (3)
	(2) edge (1);
\end{tikzpicture}
\caption{Another causal network that can exhibit Simpson's paradox. In this case, \quotes{the solution} is not to stratify on $Z$.}
\label{fig:simpsons_paradox2}
\end{figure}

@@

The structure in figure \ref{fig:dseptripple4} is referred to as a collider or v-structure. The counter-intuitive result that conditioning on $Z$ introduces dependence between $X$ and $Y$ is called the \textit{explaining away phenomena}. As an example, consider a scholarship available to female or disadvantaged students. Let $X$ be gender, $Y$ be family background and $Z$ receipt of the scholarship. There are roughly equal numbers of boys and girls in both poor and wealthy families so $X$ and $Y$ are independent. However, if we know a student is receiving a scholarship, then learning that they are male  increases the probability that they are disadvantaged.
\begin{figure}[ht]
\centering
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y| Z$}
\label{fig:dseptripple1}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$X$};
\node[main node, below =of 1](2){$Z$};
\node[main node, below =of 2](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    (2) edge node {} (3);	
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y|Z$}
\label{fig:dseptripple2}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$Y$};
\node[main node, below =of 1](2){$Z$};
\node[main node, below =of 2](3){$X$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    (2) edge node {} (3);	
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y|Z$}
\label{fig:dseptripple3}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$Z$};
\node[main node, below left =of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
        edge node {} (3);	
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y|\{\}$}
\label{fig:dseptripple4}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$Z$};
\node[main node, above left =of 1](2){$X$};
\node[main node, above right =of 1](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge node {} (1)
    (3) edge node {} (1);	
\end{tikzpicture}
\end{subfigure}
\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
\label{fig:dseptriples}
\end{figure}

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{original network, $G$}
\label{fig:mutilatedOriginal}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{original network, $G$}
\label{fig:rule2_orig}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{original network, $G$}
\label{fig:rule22_orig}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{original network, $G$}
\label{fig:rule3_original}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]
 %nodes
\node[main node](1){$Z$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, below right=of 2](4){$W$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (4)
    (4) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{augmented graph $G^{\hat{X}}$}
\label{fig:rule3_augmented}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]
 %nodes
\node[main node](1){$Z$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, below right=of 2](4){$W$};
\node[main node, above left=of 2](5){$\hat{X}$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (4)
    (4) edge (3)
    (5) edge (2);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{$G_{\overline{X}}$}
\label{fig:rule3_after}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]
 %nodes
\node[main node](1){$Z$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, below right=of 2](4){$W$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (3)
    (2) edge (4)
    (4) edge (3);
\end{tikzpicture} 
\end{subfigure}
\caption{Example application of equation \ref{eq:Do3}. $(Y \ci \hat{X}|W,Z) \implies \P{Y|do(X),W,Z} = \P{Y|W,Z}$. We have to condition on $Z$ because conditioning on $W$ blocks the path $\hat{X}-X-W-Y$ but opens $\hat{X}-X-Z-Y$.}
\label{fig:rule3}
\end{figure}

@@

\citet{Johansson2016_counterfactual_rep} identified that this is equivalent to the covariate shift problem. If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$, but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$.\footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift. Take the case where we have a binary intervention $x \in \set{0,1}$. Suppose we learn $h(1,\vb{z}) = \E{Y|x=1,\vb{z}}+g(\vb{z})$ and $h(0,\vb{z}) = \E{Y|x=0,\vb{z}}+g(\vb{z})$, then the estimated average causal effect equals the true average causal effect for any function $g$, $\E{h(1,\vb{z})-h(0,\vb{z})} = \E{Y|x=1,\vb{z}} - \E{Y|x=0,\vb{z}}$. More generally, if the goal is to select an optimal action $x^*$ from a continuous space of possible interventions we need algorithms capable of leveraging any structure in the relationship between $x$ and $y$ as well as a means of focusing the loss on regions of the sample likely to affect $x^*$.} With this connection to covariate shift in mind, let us return to regression, matching and propensity scores.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
\caption{
Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data. In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities. If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$. We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points). In both cases the linear model is a good fit to the data. However, the resulting estimate of the causal effect is very poor for the lower values of $z$. } 
\label{fig:covariate_shift_adjust}
\end{figure}
